---
title: "Hands On continious time modeling with ctsem"
author: "Valentin Kriegmair"
format: 
  revealjs:
    theme: "moon"
    slide-number: true
    transition: "none" 
---

## Outline
* Recap
* Walk-through
* Excercise
* Summary

## Recap I
What are the assumptions about time in the general cross lagged panel model (GCLPM)?

* time prgoresses in discrete steps
* equally spaced measurement occasions
* subjects are measured at the same time points

## Recap II
What limitations does discrete time modeling of longitudinal data?

* biased parameter estimates due to:
  * unequally spaced measurement waves in the panel data
  * measurement of individual subjects at each wave spread out over time

* uncomparable parameter estimates across studies unequally spaced time intervals

## Recap II
Multivariate stochastic differential equation:

\begin{gather*}
\mathrm{d} \boldsymbol{\eta}_i(t)=\left({\mathbf{A}} \boldsymbol{\eta}_i(t)+\mathbf{B} \mathbf{z}_i+\mathbf{M} \boldsymbol{\chi}_i(t)\right) \mathrm{d} t+\mathbf{G} \mathrm{d} \mathbf{W}_i(t)
\end{gather*}

## Recap II
Multivariate stochastic \color{red}{differential} equation:

\begin{gather*}
\color{red}{\mathrm{d}} \boldsymbol{\eta}_i(\color{red}{t})=\left(\mathbf{A} \boldsymbol{\eta}_i(\color{red}{t})+\mathbf{B} \mathbf{z}_i+\mathbf{M} \boldsymbol{\chi}_i(t)\right) \color{red}{\mathrm{d} t}+\mathbf{G} \color{red}{\mathrm{d}} \mathbf{W}_i(\color{red}{t})
\end{gather*}

modelling change in latent variables $\boldsymbol{\eta}_i(t)$
 as a function of time over infinitesimal small time steps $\color{red}{\mathrm{d} t}$

## Recap II
Multivariate stochastic differential equation:

\begin{gather*}
\mathrm{d} \boldsymbol{\eta}_i(t)=\left(\color{red}{\mathbf{A}} \boldsymbol{\eta}_i(t)+\mathbf{B} \mathbf{z}_i+\mathbf{M} \boldsymbol{\chi}_i(t)\right) \mathrm{d} t+\mathbf{G} \mathrm{d} \mathbf{W}_i(t)
\end{gather*}

modelling change as a function of time over infinitesimal small time steps $\mathrm{d} t$

* $\color{red}{\mathbf{A}}$: drift matrix with auto effects on the diagonal and cross effects on the off-diagonals ($\mathbf{A} \in \mathbb{R}^{v \times v}$).


## Recap II
Multivariate stochastic differential equation:

\begin{gather*}
\mathrm{d} \boldsymbol{\eta}_i(t)=\left(\mathbf{A} \boldsymbol{\eta}_i(t)+\mathbf{B} \mathbf{z}_i+\mathbf{M} \boldsymbol{\chi}_i(t)\right) \mathrm{d} t+\mathbf{G} \mathrm{d} \mathbf{W}_i(t)
\end{gather*}

modelling change as a function of time over infinitesimal small time steps $\mathrm{d} t$

* $\mathbf{A}$: drift matrix with auto effects on the diagonal and cross effects on the off-diagonals ($\mathbf{A} \in \mathbb{R}^{v \times v}$).

* $\mathbf{z_i}$: time independent predictors

* $\boldsymbol{\chi}_i(t)$: time dependent predictors

* $\mathbf{W}_i(t)$: stochastic error term (Wiener process)

* $\mathbf{G}$: diffusion matrix

## Recap II
Multivariate stochastic differential equation:
\begin{gather*}
\mathrm{d} \boldsymbol{\eta}_i(t)=\left(\mathbf{A} \boldsymbol{\eta}_i(t)+\mathbf{B} \mathbf{z}_i+\mathbf{M} \boldsymbol{\chi}_i(t)\right) \mathrm{d} t+\mathbf{G} \mathrm{d} \mathbf{W}_i(t)
\end{gather*}

measurement part:
\begin{gather*}
\mathbf{y}(t)=\boldsymbol{\Lambda} \boldsymbol{\eta}(t)+{\boldsymbol{\tau}}+\boldsymbol{\varepsilon}(t)
\end{gather*}
\begin{gather*}
\boldsymbol{\varepsilon}(t) \sim \mathrm{N}(\mathbf{0}, \mathbf{Θ})
\end{gather*}

* $\boldsymbol{\Lambda}$: factor loadings
* $\boldsymbol{\tau}$: intercepts
* $\boldsymbol{\varepsilon}(t)$: measurement error

## {background-image="figures/figure3-1.png"}
## {background-image="figures/figure3-2.png"}

## Introducing `ctsem` & `ctsemOMX`
* R-packages `ctsem` & `ctsemOMX` to handle the math
* `ctsem` interfaces with Stan
* `ctsemOMX` interfaces with OpenMx
* today we will use `ctsemOMX`
```{r}
# install.packages("ctsem")
# install.packages("ctsemOMX")
library(ctsem) # loads ctsem
library(ctsemOMX) # loads ctsem and ctsemOMX
```

## Data Structure
* `ctsemOMX` uses wide data with time intervals specified in columns:
```{r}
data("datastructure", package = "ctsem")
datastructure
```

* to convert from the typical long format use `ctLongToWide()`:
```{.r}
df_wide <- ctLongToWide(datalong = df_long, id = "id",
  time = "time", manifestNames = c("Y1", "Y2", "Y3"))
```